From heartbeat to data -- Using wearable fitness trackers as an affordable approach to assess teacher stress

### 2024-09-09 02_paper/02_study/01_fitbit paper/Submission/FitBit_submission.tex
Tabellen
Format Sachen

\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={New approaches to teachers' experience of stress: Do heart rate measurements with fitness trackers provide an efficient, inexpensive, and robust measurement method?},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}

@@

  \title{New approaches to teachers' experience of stress: Do heart rate
measurements with fitness trackers provide an efficient, inexpensive,
and robust measurement method?}
    \author[Leipzig University]{Mandy Klatt%
  \corref{cor1}%
  \fnref{1}}
   \ead{mandy.klatt@uni-leipzig.de} 
    \author[Leipzig University]{Peer Keßler%
  %
  \fnref{2}}
   \ead{Peer@example.com} 
    \author[Leipzig University, Max Planck Institute for Evolutionary
Anthropology]{Gregor Kachel%
  %
  \fnref{1}}
   \ead{gregor@example.com} 
    \author[Leipzig University]{Christin Lotz%
  %
  \fnref{1}}
   \ead{christin@example.com} 
    \author[Leipzig University]{Anne Deiglmayr%
  %
  \fnref{3}}
   \ead{anne@example.com} 
      \affiliation[Leipzig University]{
    organization={Empirische Schul- und
Unterrichtsforschung},addressline={Marschnerstr.
29},city={Leipzig},postcode={4109},state={Sachsen},country={Germany},}
    \affiliation[Max Planck Institute for Evolutionary Anthropology]{
    organization={Department},addressline={Deutscher Pl.
6},city={Leipzig},postcode={4103},state={Sachsen},country={Germany},}
    \cortext[cor1]{Corresponding author}
    \fntext[1]{ConceptualizationWriting - Original Draft
PreparationWriting - Review \& Editing}
    \fntext[2]{Writing - Original Draft PreparationWriting - Review \&
Editing}
    \fntext[3]{Supervision}

  \begin{abstract}
  One or two sentences providing a \textbf{basic introduction} to the
  field, comprehensible to a scientist in any discipline.

  Two to three sentences of \textbf{more detailed background},
  comprehensible to scientists in related disciplines.

  One sentence clearly stating the \textbf{general problem} being
  addressed by this particular study.

  One sentence summarizing the main result (with the words
  ``\textbf{here we show}'' or their equivalent).

  Two or three sentences explaining what the \textbf{main result}
  reveals in direct comparison to what was thought to be the case
  previously, or how the main result adds to previous knowledge.

  One or two sentences to put the results into a more \textbf{general
  context}.

  Two or three sentences to provide a \textbf{broader perspective},
  readily comprehensible to a scientist in any discipline.

  XXX In this proof-of-concept study, we aimed to advance the field of
  teacher stress by collecting heart rate data with wrist-worn devices
  and testing a methodology that has the potential to provide more
  insights on the non-invasive assessment of teacher stress. XXX
  \end{abstract}
    \begin{keyword}
    heart rate \sep photoplethysmography \sep wearable electronic
device \sep 
    teaching
  \end{keyword}

@@

In education, there's interest in identifying reliable indicators for
teacher stress and burnout \citep{fisher2011, junker2021}. Previous
research often relied on self-report questionnaires
\citep{chaplain2008, liu2020}, which may be prone to biases like social
desirability \citep{razavi2001self} or recall errors
\citep{van2016accuracy}. To overcome these limitations, ambulatory
assessment methods are recommended
\citep{trull2013ambulatory, wettstein2020ambulatory}, such as measuring
physiological parameters like heart rate (HR), offering objective
insights into teachers' stress levels without disrupting their teaching
\citep{donker2018, runge2020}.

Current studies on teachers' HR in teaching settings often rely on
expensive and invasive electrocardiographs
\citep{sperka1995, scheuch1997psychophysische, donker2018, junker2021, huang2022class},
showing that teacher-centered activities and common stressors lead to
increased HR. However, using affordable and non-invasive instruments
like wrist-worn fitness trackers \citep{ferguson2015} could enhance HR
recording in educational contexts. Unlike clinical devices, fitness
trackers could offer continuous and less intrusive data collection over
time \citep{godfrey2018z}, aligning with the increasing popularity and
acceptance of wearables among the general population
\citep{peng2022acceptance}. These devices, equipped with biosensors,
provide users with physiological and behavioral data, offering an
accessible means for monitoring physical activity and health in daily
life.

It can be assumed that teachers also wear personal, private fitness
trackers, generating recordings of physiological data that could be a
very interesting resource for research on teacher stress. While the use
of wearable technology in the field has been explored in various domains
{[}\citet{hughes2023wearable}; @ adesida2019exploring;
helmer2009smart{]}, research is sparse in educational contexts
\citep{de2017towards}. Although some studies investigated how wearables
can help teachers monitor student activity \citep{quintana2016keeping},
there is a notable research gap regarding teachers' use of wrist-worn
wearables. Given the high stress levels in the teaching profession
\citep{johnson2005experience}, fitness trackers could be a valuable tool
for analyzing HR and the factors contributing to stress. One of the
reasons for teachers' augmented stress is that they are confronted with
a multitude of demands in their everyday work, some of which exceed
their resources and therefore make it difficult to cope with immediate
stressors such as classroom disruptions \citep{montgomery2005meta}.
However, the extent of the strain depends on the subjective appraisal of
a stressor, which involves considerations about available resources to
deal with it \citep{kyriacou2001}. It is, therefore, particularly
important for teachers to have sufficient personal and professional
resources at their disposal \citep{cramer2018belastung}. Classroom
disruptions, for example, are one of the major stressors in teachers'
daily work \citep{boyle1995structural, aloe2014multivariate} and
professional knowledge about effective classroom management reduces the
risk of teacher stress \citep{klusmann2012berufliche}. Teachers'
characteristics such as professional experience in turn have an impact
on the development of classroom management skills and thus on the
appraisal processes, as these skills develop during professional
experience \citep{ophardt2017klassenmanagement, wolff2015keeping}.

To better understand how stressors like classroom disruptions affect
teachers and their stress responses, data from wrist-worn fitness
trackers could be used to monitor physiological parameters before,
during, and after teaching sessions \citep{wettstein2021}. This study
explored the use of wrist-based fitness trackers as a tool to monitor
teachers' stress during different phases of a micro-teaching unit during
which teachers had to deal with classroom disruptions. Physiological
data were triangulated with teachers' appraisal of classroom
disruptions, and their teaching experience. The physiological indicator
employed in this study was teachers' HR, which can be readily recorded
by any fitness tracker. Teachers' use of wrist-worn fitness trackers in
educational research holds transformative potential as obtained
fine-grained data and underscores the critical need to monitor teachers'
health as they navigate the stressful landscape of the classroom.

\subsection{Fitness Trackers as a Method to Assess HR as an Indicator of
Stress}\label{fitness-trackers-as-a-method-to-assess-hr-as-an-indicator-of-stress}

Wearables, defined as electronic devices worn directly on the body or
integrated into clothing or accessories, serve as versatile solutions
\citep{godfrey2018z}, gathering data like location, movements, and vital
signs \citep{cheng2017underlying}. Fitness trackers, a popular wearable
type \citep{park2020user}, offer insights into physical activity and
cardiovascular health, supporting personalized fitness goals
\citep{nuss2021effects} and stress management \citep{hao2018chrv}. Their
affordability and ease of use make them ideal not only for various
domains including healthcare, entertainment, and fitness
{[}sinha2019taxonomy{]} but also in education as they offer added
benefits for formal and informal learning environments for both students
and teachers \citep{de2017towards}. Yet, few studies focus on their
significance for teachers.

One important health parameter assessed by nearly all wrist wearables is
HR \citep{scalise2018wearables}. HR indicates the number of heartbeats
within one minute and is typically expressed as beats per minute (BPM)
\citep{hottenrott2007}. At rest, the average HR of adults typically
ranges from 60 to 80 BPM \citep{sammito2015guideline}. HR can be
detected and measured using sensors based on electrocardiogram (ECG) or
phonocardiogram (PCG) \citep{mukhopadhyay2017wearable}. Another
uncomplicated and inexpensive technique to measure HR via fitness
trackers is photoplethysmography (PPG) \citep{castaneda2018review}. This
optical method assesses HR by flashing green or red lights to measure
changes in blood volume \citep{allen2007photoplethysmography}.\\
Physiologically, HR is regulated and influenced on short-time intervals
by the sympathetic and the parasympathetic nervous system
\citep{pham2021}. An increase in the activity of the sympathetic system
results in HR being speeded up (``fight or flight'')
\citep{taelman2009influence}. In contrast, increased activity of the
parasympathetic has the effect of slowing down the HR (``rest and
digest'') \citep{battipaglia2015}. Therefore, an increase in HR can be
regarded as an indicator of increasing stress, and a decrease as an
indicator of decreasing stress \citep{kyriacou1978}.

\subsection{Teacher Stress and Important
Resources}\label{teacher-stress-and-important-resources}

The teaching profession is one of the most stressful professions
compared to other occupational groups, facing a variety of stressors
during everyday work \citep{smith2000, herman2020, schult2014belastet}.
According to \citet{kyriacou1978}, teacher stress can be defined as a
negative affective response, often accompanied by physiological changes
such as increased HR, triggered by job-related demands and perceived as
threatening to one's self-esteem or well-being. Coping mechanisms help
to reduce the perceived threat.

This definition of teacher stress is based on the transactional stress
model by Lazarus and colleagues
\citep{lazarus1981stressbezogene, lazarus1984stress}, which was modified
and tailored to the teaching-learning environment by
\citet{kyriacou1978}.

In general, the transactional stress model \citep{lazarus1990theory}
highlights the interaction between an individual and the environment,
whereby stress refers to any event that exceeds a person's adaptive
resources. It has been shown that there are important connections
between stressors and resources experienced by teachers on the one hand,
and stress-induced health issues on the other hand
\citep{krause2013messung}. As we are interested in how specific
classroom events affect teacher stress, we adapted the transactional
stress model to that type of situation, based on a representation of the
model first proposed by {[}van2006stress{]}; this working model is
depicted in Fig. 1.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1\textwidth]{images/Model_Teacher_Stress_adapted_new.pdf}
  \caption{A model of teacher stress (adapted from van Dick 2006, p.37, modified by the author)}
  \label{A model of teacher stress (adapted from van Dick 2006, p.37, modified by the author)}
\end{figure}

When potential stressors (e.g., classroom disruptions) occur during
teaching (1), teachers subjectively appraise how disruptive the event is
(2). If potential stressors are judged as threatening, i.e., as actual
stressors (3), teachers consider whether they have sufficient resources
for coping with the stressor (4). Based on this evaluation of resources
and their characteristics such as professional experience, teachers try
to manage classroom disruptions, by utilizing classroom management
strategies (5). In cases where coping fails, stress ensues, often
accompanied by physiological reactions like increased HR (6). Following
resource appraisal and coping outcomes, the stressor is re-evaluated
(7).

The appraisal processes of resources play a crucial role as teachers'
assessments of resources and coping mechanisms, both internal (e.g.,
classroom management skills) and external (e.g., supportive colleagues),
significantly impact their stress levels. When resources are lacking and
coping fails, negative consequences like burnout and high turnover can
arise \citep{jalongo2006, unterbrink2007, aloe2014}. This highlights the
importance of teachers' appraisal of their professional competencies,
including their ability to manage classroom disruptions effectively
\citep{klieme2008concept, konig2016teacher}.

As shown in Fig. 1, both primary and secondary appraisals are influenced
by teachers' characteristics, e.g., their professional experience,
shaping their classroom management skills. As experience grows, teachers
develop cognitive scripts for managing classroom events resulting in
more complex classroom management skills \citep{wolff2021classroom}.
Especially beginning teachers face considerable stress and often feel
overwhelmed by the demands of teaching
{[}\citet{ophardt2017klassenmanagement}; \citet{wolff2015keeping}; @
klusmann2012berufliche{]}, with many leaving the profession within the
first five years \citep{ingersoll2003}. Research shows less experienced
teachers are more susceptible to burnout, underscoring the importance of
experience and job satisfaction in predicting teacher stress
\citep{fisher2011}.

\subsection{HR in Teaching-Learning
Contexts}\label{hr-in-teaching-learning-contexts}

ECG studies have revealed that HR can be used to map changes in HR
during teaching onto stressors experienced by teachers. For example, HR
increased during teacher-centered activities when teachers had to take a
leading position in the student-teacher interaction
\citep{sperka1995, scheuch1997psychophysische, donker2018, junker2021}.
\citet{sperka1995} for example recorded the HR of 16 pre-service
teachers during their first lesson. The results showed significantly
increased psychophysiological activation in terms of an increased HR
during teaching. The activation effect was particularly prominent at the
beginning of the lesson and decreased over the course of the lesson. The
authors interpret this result as showing how pre-service teachers'
active coping processes, i.e., the active management of the interaction
with the students, helped the teachers regulate their HR downwards.
Other ECG studies identified typical stressors predicting increased HR
values, such as class size \citep{huang2022class}, or low student
engagement and motivation \citep{junker2021}. For example,
\citet{junker2021} recorded the HR of 40 teachers using an ambulatory
monitoring system during a real classroom lesson. They provided evidence
that teacher stress caused by stressors such as low student engagement,
i.e., students displaying a lack of motivation or limited interest in
their tasks or classroom activities, or teacher-centered activities,
i.e., classroom activities primarily focused on the teacher's actions.,
leads to an increase in HR.

In addition to ECG studies, there are a few studies that used wrist-worn
fitness trackers to investigate HR trends in teaching-learning
situations \citep{Darnell2019, chalmers2021}. \citet{Darnell2019} for
example measured the HR of 15 medical college students using wrist-worn
devices during lectures. The analysis revealed a constant decrease in HR
from the beginning to the end of a lecture, whereas the HR peak was
reached during active learning sessions. \citet{chalmers2021} examined
the usability of the average HR, measured with a Fitbit fitness tracker,
to identify physiological changes during stress-inducing tasks in a
study with a total of 60 participants. The average HR increased
significantly between the resting and stress phases. Even though the
participants in these studies were learners, not teachers, the results
are relevant to the study of teacher stress using wearable devices,
because the studies showed that a) HR can be effectively recorded using
fitness trackers during a whole learning unit, and b) HR changes are in
line with the occurrence of activating or stress-inducing tasks.

So far, to the best of our knowledge, only one study has directly
assessed teachers' HR using a wrist-worn fitness tracker during teaching
: \citet{runge2020} used a Fitbit fitness tracker to assess HR as an
indicator of stress in N=4 teachers. They used the fitness trackers'
recordings to create a profile for each teacher (to differentiate
between teachers reporting higher or lower levels of stress .) In
particular, it was found that the combination of a high number of steps,
a high HR, and short sleep was an indicator of stress. It should be
noted that the generalizability of the results is limited due to the
small sample size.

In summary, previous studies have revealed that teachers' (and
students') HR changes, depending on the activity and stressors they
experience, whereby teacher-centered phases , in particular, led to an
increase in HR
\citep{sperka1995, scheuch1997psychophysische, donker2018, junker2021}.
Furthermore, it could be shown that HR as an indicator of stress can be
assessed using low-cost, non-intrusive fitness trackers and that HR
increases in activating phases and even before stress occurs
\citep{Darnell2019, chalmers2021}. However, studies collecting data from
teacher-worn fitness trackers in a large enough sample to explore links
with factors such as subjective stressor appraisal, or effects of
teaching experience, are still lacking. Our study aims to close this
research gap.

@@

The data analyzed for the present study were obtained from teachers and
student teachers who participated in a lab study, as part of a larger
project targeting the development of classroom management in teachers.

As part of the larger project, participants came to the lab
individually, and each taught a 15-minute, self-prepared micro-teaching
unit to a ``class'' of three actors (trained student assistants). These
actors performed nine, possibly disruptive, classroom events. The actors
received standardized instructions on a screen (only visible to them,
not to the participant) to perform a classroom event every one and a
half minutes, and they performed the same scripted disruptions for all
participants. While teaching, participants wore eye-tracking glasses,
and additionally, their lessons were recorded by cameras.

The micro-teaching unit, with its unfamiliar setting and the scripted
disruptions of participants' teaching flow, was potentially stressful.
Thus, we were particularly interested in the development of
participants' HR before, during, and after this micro-teaching unit. We
recorded HR data in five phases, with a total duration of approximately
two hours: In the \emph{pre-teaching phase}, participants were welcomed,
prepared for the following micro-teaching unit, and familiarized with
the setting. In the \emph{teaching phase}, participants taught the
micro-teaching unit and experienced the classroom disruptions. In the
\emph{post-teaching phase}, participants answered several questionnaires
in a comfortable seated position. Next, in the \emph{interview phase},
they watched the video of their 15-minute unit, rated the disruptive
classroom events, and answered open questions. Finally, in the \emph{end
phase}, participants answered another questionnaire. These conditions
were identical for all participants. During the entire study,
participants wore a fitness tracker on their wrist.

The goals of the present study were twofold:

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\tightlist
\item
  The first research goal was to investigate whether HR measures
  assessed by wrist-based fitness trackers are a suitable and effective
  method for mapping teachers' HR over the course of a five-phase lab
  study, including the time before, during, and after a potentially
  stressful micro-teaching unit.
\end{enumerate}

First, we expected the participants to show an initial increase in their
HR, followed by a peak during the teaching phase and a decrease for the
remaining phases. In addition, we examined whether z-standardization of
the participants' mean HR could serve as a useful method to account for
individual differences in baseline HR. We expected to observe the same
trends in both standardized and non-standardized mean HR values.

Second, five representative 10-minute intervals were selected from the
five phases (see section \#\#Measures for a more detailed description of
the intervals), which served as the basis for the data analysis for our
hypothesis: pre-teaching interval (\(I_1\)), teaching interval
(\(I_2\)), post-teaching interval (\(I_3\)), interview interval
(\(I_4\)), end interval (\(I_5\)). We examined the levels of and the
changes in HR during these intervals. We assumed the highest HR level in
the teaching interval (\(I_2\)) and lower levels in all other intervals
(\textbf{Hypothesis 1a}). Further, we expected an increase in
participants' HR while they were preparing for teaching during the
pre-teaching interval (\(I_1\)), and we expected a decrease in
participants' HR during all of the following intervals, because of
habituating to (\(I_2\)) and recovering from (\(I_3\)-\(I_5\)) the
stressful teaching phase (\textbf{Hypothesis 1b}).

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\setcounter{enumi}{1}
\tightlist
\item
  The second research goal was to examine whether variance in HR
  measures could be explained by participants' teaching experience
  (because, presumable, more experienced teachers might have better
  classroom management strategies, and thus better resources for
  coping), and/or by their self-reported subjective appraisals of
  classroom events (specifically, how disruptive it was, and how
  confident they felt in their coping).
\end{enumerate}

>>>>

\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={From heartbeat to data -- Using wearable fitness trackers as an affordable approach to assess teacher stress},
            colorlinks=false,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}

@@

  \title{From heartbeat to data -- Using wearable fitness trackers as an
affordable approach to assess teacher stress}
    \author[Leipzig University]{Mandy Klatt%
  \corref{cor1}%
  \fnref{1}}
   \ead{mandy.klatt@uni-leipzig.de} 
    \author[Leipzig University]{Peer Keßler%
  %
  \fnref{2}}
   \ead{Peer@example.com} 
    \author[Leipzig University, Max Planck Institute for Evolutionary
Anthropology]{Gregor Kachel%
  %
  \fnref{1}}
   \ead{gregor@example.com} 
    \author[Leipzig University]{Christin Lotz%
  %
  \fnref{1}}
   \ead{christin@example.com} 
    \author[Leipzig University]{Anne Deiglmayr%
  %
  \fnref{3}}
   \ead{anne@example.com} 
      \affiliation[Leipzig University]{
    organization={Empirische Schul- und
Unterrichtsforschung},addressline={Marschnerstr.
29},city={Leipzig},postcode={4109},state={Sachsen},country={Germany},}
    \affiliation[Max Planck Institute for Evolutionary Anthropology]{
    organization={Department},addressline={Deutscher Pl.
6},city={Leipzig},postcode={4103},state={Sachsen},country={Germany},}
    \cortext[cor1]{Corresponding author}
    \fntext[1]{ConceptualizationWriting - Original Draft
PreparationWriting - Review \& Editing}
    \fntext[2]{Writing - Original Draft PreparationWriting - Review \&
Editing}
    \fntext[3]{Supervision}

  \begin{abstract}
  Past research on physiological indicators of teacher stress often had
  to rely on expensive and obtrusive assessment methods. Modern fitness
  trackers represent a non-invasive and convenient alternative. This
  study investigated the use of wrist-worn fitness trackers to assess
  teacher heart rate (HR) as an indicator of stress during teaching. In
  a laboratory study, we used a Fitbit® fitness tracker to assess
  teachers´ HR before, during, and after a potentially stressful
  micro-teaching session. Our results demonstrate that the fitness
  tracker was indeed useful for mapping teachers' stress, with the data
  showing how teachers' HR increased before, peaked during, and
  progressively decreased after the micro-teaching session. Moreover, we
  related the fitness tracker data to retrospective teacher
  self-reports. We found that teachers' subjective stress appraisals,
  together with their teaching experience, explained only small amounts
  of variance in HR data. We discuss the potential of fitness trackers
  as an affordable and ubiquitous assessment tool for research on
  teacher stress in the classroom and provide advice for practical
  implementation.
  \end{abstract}
    \begin{keyword}
    teacher stress \sep fitness tracker \sep heart rate \sep classroom
disruptions \sep wearable technology \sep 
    physiological stress measurement
  \end{keyword}

@@

The teaching profession is one of the most stressful professions, with
teachers facing a host of stressors during their everyday work
\citep{smith2000, herman2020, schult2014belastet}. To better understand
mechanisms in teacher stress, there is a growing research interest in
physiological measures such as heart rate (HR) as online measures of
teachers' stress during teaching
\citep{karner2021teachers, wettstein2020ambulatory}. For example, it has
been shown that teacher-centered activities and typical
classroom-related stressors increase teacher HR during teaching
activities
\citep{sperka1995, scheuch1997psychophysische, donker2018, junker2021, huang2022class}.
However, previous studies have often relied on expensive and obtrusive
electrocardiographs (ECG). Modern fitness trackers represent a
non-invasive and convenient alternative. \citep{ferguson2015}

Classroom disruptions are a major stressor in teachers' daily work
\citep{boyle1995structural, aloe2014multivariate}, and learning how to
deal with them is an important aspect of professional expertise
\citep{wolff2015keeping}. According to \citet{lazarus1990theory}
transactional model of stress and coping, the experience of stress in
response to stressors such as classroom disruptions depends on the
teacher´s subjective appraisal, which, in turn, depends on their coping
resources, such as their professional knowledge. The resulting stress
response has a psychological, physiological, or behavioral dimension
\citep{kyriacou1978}. Therefore, in order to better understand how
classroom stressors affect teachers' stress response, subjective
self-reports should be accompanied by objective, physiological measures
\citep{wettstein2021}. Teachers' use of wrist-worn fitness trackers in
educational research provides fine-grained, in vivo data, allowing
researchers as well as teachers themselves to monitor their
physiological stress response continuously during teaching, across
settings, and at low costs. Being able to monitor, and eventually
counteract, teacher stress levels appears particularly relevant given
the profession's generally high stress levels and associated negative
health effects \citep{johnson2005experience, montgomery2005meta}. To
harness this potential, the present study explored the use of
wrist-based fitness trackers as a tool to assess teachers' HR, as an
indicator of stress, before, during, and after a teaching session during
which typical, potentially stressful, classroom disruptions occurred.
Further, we explored to what extent teachers' subjective appraisals of
classroom disruptions and their teaching experience predicted teacher
stress as assessed by the fitness tracker.

\subsection{Fitness trackers as a ubiquitous, low-cost tool for
assessing physiological stress
responses}\label{fitness-trackers-as-a-ubiquitous-low-cost-tool-for-assessing-physiological-stress-responses}

Fitness trackers provide data on physical activity and cardiovascular
parameters such as HR, supporting personalized fitness goals
\citep{nuss2021effects} and stress management \citep{hao2018chrv}. They
can be used as ubiquitous, low-cost, and unintrusive data collection
instruments \citep{godfrey2018z}, and their wide-spread use and everyday
availability align with the increasing popularity and acceptance of
wearables among the general population \citep{peng2022acceptance}. In
contrast to self-reported questionnaires on stress
\citep{chaplain2008, liu2020} that are prone to biases like social
desirability \citep{razavi2001self} or recall errors
\citep{van2016accuracy}, fitness trackers, as ambulatory assessment
methods \citep{trull2013ambulatory, wettstein2020ambulatory}, offer more
objective insights into teachers' stress levels by monitoring teachers'
physiological stress responses without disrupting teaching
\citep{donker2018, runge2020}.

One important health parameter assessed by nearly all wrist-worn fitness
trackers is heart rate \citep{scalise2018wearables}. HR indicates the
number of heartbeats within one minute and is typically expressed as
beats per minute (BPM)
\citep{berntson2007cardiovascular, hottenrott2007}. At rest, the average
HR of adults typically ranges between 60 and 80 BPM
\citep{sammito2015guideline}. HR can be detected and measured in
different ways using sensors, such as electrocardiography (ECG) or
photoplethysmography (PPG) \citep{mukhopadhyay2017wearable}. While ECG
sensors offer precise measurements by detecting the heart's electrical
activity, their intrusive nature and requirement of direct skin contact
may limit their suitability \citep{kranjec2014non}, particularly in
educational settings. In contrast, photoplethysmography (PPG) is a
rather uncomplicated and inexpensive technique to measure HR, commonly
found in commercially available fitness trackers
\citep{castaneda2018review}. This optical method assesses HR by flashing
green or red lights to measure changes in blood volume in the
capillaries of the skin \citep{allen2007photoplethysmography}.

Physiologically, HR is regulated by the sympathetic and parasympathetic
nervous systems \citep{pham2021}. An increase in sympathetic activity
results in HR being sped up (``fight or flight'' response)
\citep{taelman2009influence}; whereas an increase in parasympathetic
activity results in HR being slowed down (``rest and digest'' response)
\citep{battipaglia2015}. Stress or mental and physical strain directly
increases HR \citep{custodis2014heart, sachs2014}. Therefore, an
increase in HR can be regarded as an indicator of increasing stress, and
a decrease as an indicator of relaxation and ease \citep{kyriacou1978}.
Thus, fitness trackers offer low-cost and unobtrusive access to
psychological stress data.

\subsection{HR in teaching-learning
contexts}\label{hr-in-teaching-learning-contexts}

Prior research, typically using ECG methods, has shown that changes in
teachers' HR can be mapped onto stressors they experience during
teaching. For example, teachers' HR tends to increase when teachers take
an exposed position in student-teacher interaction
\citep{sperka1995, scheuch1997psychophysische, donker2018, junker2021}.
\citet{sperka1995} for example recorded the HR of 16 pre-service
teachers during their first lesson and showed that teachers' HR
increased significantly during teaching. The activation was particularly
prominent at the beginning of the lesson and decreased over the course
of the lesson. The authors suggested that pre-service teachers'
proactive coping strategies, such as actively managing student
interactions, helped lower their HR levels. Other ECG studies identified
typical stressors predicting increases in HR, such as class size
\citep{huang2022class}, or low student engagement and motivation
\citep{junker2021}. \citet{junker2021} recorded the HR of 40 teachers
during a real classroom lesson. Again, teacher stress, induced by
factors such as low student engagement (e.g., lack of motivation or
interest in tasks) or teacher-centered activities (e.g., teacher-focused
classroom activities), resulted in elevated HR.

More recent studies have begun using wrist-worn fitness trackers to
investigate HR trends in instructional settings
\citep{Darnell2019, chalmers2021}. \citet{Darnell2019} measured the HR
of 15 medical college students listening to lectures, using wrist-worn
devices. The analysis revealed a constant decrease in HR throughout the
lecture, with HR peaks during more interactive learning phases.
\citet{chalmers2021} used HR data from a fitness tracker to identify
physiological changes during stress-inducing tasks (i.e., the Trier
Social Stress Test; \citet{kirschbaum1993trier}{]}. Average HR increased
significantly from the resting to the stress inducing phases. Even
though the participants of these previous studies
\citep{Darnell2019, chalmers2021} were not teachers but learners, the
results are relevant for studying teacher stress as they demonstrated
that HR can be effectively recorded using fitness trackers over the
course of a whole learning unit, and HR changes are in line with the
occurrence of activating or stress-inducing tasks.

To the best of our knowledge, only one study has directly assessed
teachers' HR during teaching using a wrist-worn fitness tracker:
\citet{runge2020} assessed HR as an indicator of stress in four
in-service teachers during authentic lessons. They used fitness
trackers' recordings to create a profile for each teacher, with the aim
of differentiating between teachers reporting higher vs.~lower levels of
stress. It was found that the combination of a high HR, a high number of
steps, and short sleep duration was characteristic for teachers
reporting high stress levels. However, it should be noted that the
generalizability of these results is limited due to the small sample
size.

In summary, previous studies have revealed that teachers' (and
students') HR changes depend on their activities and the stressors they
experience, with an increase in HR already before the expected stressors
occur, and with peaks in activating phases
\citep{Darnell2019, chalmers2021}. For teachers, teacher-centered phases
led to an increase in HR
\citep{sperka1995, scheuch1997psychophysische, donker2018, junker2021}.
However, there is a lack of studies using teacher-worn fitness trackers
in larger samples, exploring the feasibility of this convenient tool for
researching links between teachers' HR and stress inducing events and
mechanisms.

\subsection{A model of teacher stress}\label{a-model-of-teacher-stress}

According to \citet{kyriacou1978}, teacher stress can be defined as a
negative affective response, typically accompanied by physiological
changes such as increased HR, triggered by job-related demands, and
perceived as threatening to one's self-esteem or well-being. Coping
mechanisms help to reduce the perceived threat. Kyriacou's definition of
teacher stress (see \citet{kyriacou1978} and, for a more recent
adaptation, \citet{van2006stress}) is based on the transactional stress
model by Lazarus and colleagues {[}lazarus1966psychological;
\citet{lazarus1990theory}{]}, which highlights the interaction between
an individual and the environment, whereby stress refers to a person's
subjective reaction to an event (a stressor) that exceeds the person's
adaptive resources.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1\textwidth]{images/Model_Teacher_Stress_adapted_new.pdf}
  \caption{A model of teacher stress (adapted from van Dick 2006, p.37, modified by the authors)}
  \label{A model of teacher stress (adapted from van Dick 2006, p.37, modified by the authors)}
\end{figure}

Fig. 1 shows, in a simplified way, how classroom events affect teachers'
stress level, according to the adaptation of the Lazarus model to
teacher stress proposed by \citet{van2006stress}: When potential
stressors (e.g., classroom disruptions) occur during teaching (1),
teachers intuitively judge how disruptive the event is in a primary
appraisal (2). If potential stressors are judged as threatening, i.e.,
as actual stressors (3), teachers consider whether they have sufficient
resources for coping with the stressors (4). Teachers utilize these
resources in trying to cope with the stressors, e.g., by employing
classroom management strategies (5). In cases when coping fails, stress
ensues, often accompanied by physiological reactions like increased HR
(6). As part of the coping process, and dependent on its outcomes,
teachers re-evaluate the stressor (7).

As shown in Fig. 1, teachers' primary and secondary appraisal as well as
coping attempts are influenced by professional experience. As
professional experience grows, teachers develop cognitive scripts for
managing classroom events, resulting in more complex and problem-focused
classroom management skills \citep{wolff2021classroom}, and thus more
effective coping and less stress. Empirically, classroom management
skills and problem-focused coping styles are linked to fewer instances
of emotional exhaustion \citep{maslach2001job, clunies2008self}. Novices
in the teaching profession, on the other hand, face considerable stress
and often feel overwhelmed by the demands of teaching
\citep{ophardt2017klassenmanagement, wolff2015keeping, klusmann2012berufliche}
with many leaving the profession within the first five years
\citep{ingersoll2003}. Accordingly, when resources are lacking and
coping fails, negative consequences for health (e.g., burnout) and for
work (e.g., high turnover rates) can arise
\citep{jalongo2006, unterbrink2007, aloe2014multivariate}, highlighting
the importance of professional expertise for managing teacher stress
\citep{fisher2011}.

@@

The present study aimed to explore the relations between teachers' HR
response, and their subjective appraisals of stress during a
micro-teaching unit, and to relate their self-reported appraisals and
physiological stress responses to their teaching experience. We analyzed
data from in-service and pre-service teachers who participated in a
laboratory study as part of a larger project targeting the development
of classroom management skills. Participants came to the lab
individually and taught a short lesson to a class of three actors (i.e.,
trained student assistants) who performed several typical and possibly
disruptive classroom events. The micro-teaching unit was thus
potentially stressful for the participants. The aims of the present
study were twofold:

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\tightlist
\item
  The first research goal was to investigate whether HR measures
  assessed by a wrist-based fitness tracker were a suitable and
  effective method for mapping teachers' HR over the course of the lab
  study, with a total duration of approximately 2 hours, including
  phases before, during, and after the stressful micro-teaching unit.
\end{enumerate}

Looking at HR measures globally, we expected the participants to show an
initial increase in their HR, followed by a peak during the
micro-teaching unit and a decrease for the remaining phases. In
addition, we examined whether z-standardization of the participants' HR
could serve as a useful method to account for individual differences in
baseline HR: We expected to observe the same trends in both standardized
and non-standardized HR values.

In addition, we selected five representative 10-minute intervals from
the five phases of the lab study (see Figure 2) in order to test the
hypotheses that, regarding HR levels, teachers' HR would be the highest
during the micro-teaching unit, compared to all other phases
(\textbf{Hypothesis 1a}), and, regarding HR slopes, that teachers' HR
would increase while they were preparing for teaching (pre-teaching
interval), but decrease in all of the following intervals, i.e.~when
they were habituating to and recovering from the stressful
micro-teaching unit (\textbf{Hypothesis 1b}).

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\setcounter{enumi}{1}
\tightlist
\item
  We further explored whether teaching experience made a difference in
  how teachers' HR reacted to the classroom disruptions. We expected
  more experienced teachers to be less stressed by the classroom events
  (\textbf{Hypothesis 2a}). In addition, we examined the relations
  between teachers' subjective appraisals of the classroom events
  (specifically, the disruptiveness of the events, and their confidence
  in dealing with them) and teachers' HR level, beyond the explanatory
  power of teaching experience. We expected higher HR levels for
  teachers who felt more disrupted, regardless of their teaching
  experience (\textbf{Hypotheses 2b}), and lower HR levels for teachers
  who felt more confident, regardless of teaching experience
  (\textbf{Hypothesis 2c}). We hypothesized that each of the three
  predictors (\emph{teaching experience, disruption appraisal,
  confidence appraisal}) uniquely contributed to explaining variance in
  teachers' HR levels (\textbf{Hypothesis 2d}). In addition, we
  exploratively ran analogous analyses for the \emph{changes} in HR
  (i.e., slopes).
\end{enumerate}
